Combining global tree cover loss data with national historical forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar

### 2017-03-01 deforestmap.Rmd
Update methods part

In this study, we present a simple approach combining maps from @Harper2007 and products from @Hansen2013 to derive annual and cloud-free forest-cover change map for Madagascar for the period 2000-2014. Basically, we used the forest map provided by @Harper2007 for the year 2000 (defining the land-use) with the tree cover loss product provided by @Hansen2013 that we applied only inside forest areas identified by @Harper2007. Following the work done by @Harper2007, we studied in details the change in deforestation rates and the evolution of forest fragmentation on six decades from _c._ 1953 to 2014. We showed that despite all the conservation efforts done in Madagascar since the mid 1980s, deforestation has increased in speed since 2010 and that fragmentation has dramatically increased leading to few remaining intact forest landscapes. The approach described in this study could be easilly extended to all tropical countries with a forest map obtained at one date between 2000 and 2014, and could be repeated in the future to obtain future forest-cover change maps, as and when the @Hansen2013 product is updated.

@@

### 2.1 Historical deforestation maps

We derived 30m cloud-free forest maps for Madagascar for the years 1990, 2000, 2010 and 2014 combining the 1990 and 2000 forest maps by @Harper2007 and the 2000-2014 tree cover and tree cover loss maps by @Hansen2013. On the 2000 Harper's map, ~200,000 ha clouds are present over the ~4.2 million ha moist forest ecoregion (4.8% of clouds). To remove these clouds, we used the 2000 cloud-free tree cover map provided by @Hansen2013 (also at 30~m resolution) and we chose a threshold of 75% of tree cover [@Achard2014] to decide whether to replace cloud pixels with forest or non-forest pixels. We thus obtained a cloud-free year 2000 forest map for Madagascar. From this map, using tree cover loss data from 2000 to 2014 provided by @Hansen2013, we obtained cloud-free forest maps in 2010 and 2014 at 30~m resolution. Finally, we corrected the cloudy 1990 Harper forest map assuming that if forest was present in 2000, it was also present in 1990. Doing so, all clouds were removed and we obtained a cloud-free 1990 forest map for Madagascar. 
For all forest cover maps, we also removed the isolated non-forest pixels if all the height surrounding pixels were covered by forest, assuming that small-scaled deforestation on a 30m square pixel inside the forest is unlikely. This was important in order not to over-estimate forest fragmentation.  
These maps are freely available at http:/bioscenemada.cirad.fr/forestmaps. On the basis of the forest-cover maps, we computed the mean annual deforestation intensity (in hectares of forest and in percentage) on six decades for 5 time periods from _c._ 1953 to 2014 (_c._ 1953-_c._ 1973-1990-2000-2010-2014).

### 2.2 Comparison with other historical deforestation maps

We estimated the accuracy of our approach combining Harper's and Hansen's products to derive cloud-free forest-cover change maps after the year 2000 in Madagascar. To do so, we used a forest-cover change map obtained from a regional study on the humid ecoregion on the period 2005-2013 [@Rakotomalala2015]. @Rakotomalala2015 performed a supervised classification of Landsat images at multiple dates using the Random Forest classifier [@Grinand2013]. The map by @Rakotomalala2015 is considered rather accurate on the moist ecoregion with an external validation on nearly 10,854 visual interpretation points leading to rather high Kappa and overall-accuracy values (0.79 and 0.89 respectively). Following our methodology, we produced a deforestation map on the same period of time (2005-2013) and compared the forest-cover (in ha) and mean annual deforestation rates (in ha and % per year) between the two maps.

@@

We report an analysis of the evolution of forest fragmentation based on the forest-cover maps for Madagascar for the years _c._ 1953, _c._ 1973, 1990, 2000, 2010 and 2014. We used the method developed by @Riitters2000 which uses a moving window to characterize the fragmentation around each forested pixel. Computation were done using function `r.forestfrag` from GRASS GIS software [@Neteler2008]. Six categories of fragmentation are identified from the amount of forest and its occurrence as adjacent forest pixels: interior, perforated, edge, transitional, patch, and undetermined. We used a moving window of 7x7 pixels (4.41 km<sup>2</sup>). Using this window size, forest edge had a width of about 90 m [@Riitters2000]. The interior category can be interpreted as the most intact forest (sensus @Potapov2017). We reported the area of forest in each fragmentation category for the six years and analysed the evolution of the fragmentation with time.

@@

Difference might be due to post-processing of @Rakotomalala2015 map.

@@

```{r defor_comp, results="asis"}
tab.3 <- defor.comp
tab.3$ForestType <- c("Total",rep("",3),"Moist",rep(" ",3),"Dry",rep(" ",3),"Spiny",rep(" ",3),"Mangroves",rep(" ",3))
col.names <- c("Forest type","Source","1953-1973","1973-1990","1990-2000","2000-2005","2005-2010","2010-2013")
kable(tab.2,col.names=col.names,align=c("l","l",rep("r",6)))
```

@@

>>>>

In this study, we present a simple approach combining maps from @Harper2007 and products from @Hansen2013 to derive annual and cloud-free forest-cover change map for Madagascar for the period 2000-2014. Basically, we used the natural forest-cover map provided by @Harper2007 for the year 2000 (defining the land-use) with the tree cover loss product provided by @Hansen2013 that we applied only inside forest areas identified by @Harper2007. Following the work done by @Harper2007, we studied in details the change in deforestation rates and the evolution of forest fragmentation on six decades from _c._ 1953 to 2014. We showed that despite all the conservation efforts done in Madagascar since the mid 1980s, deforestation has increased in speed since 2010 and that fragmentation has dramatically increased leading to few remaining intact forest landscapes. The approach described in this study could be easilly extended to all tropical countries with a forest map obtained at one date between 2000 and 2014, and could be repeated in the future to obtain future forest-cover change maps, as and when the @Hansen2013 product is updated.

@@

### 2.1 Forest-cover maps for Madagascar from _c._ 1953 to 2014

We derived 30m cloud-free forest maps for Madagascar for the years 1990, 2000, 2010 and 2014 combining the 1990 and 2000 forest maps by @Harper2007 and the 2000-2014 tree cover and tree cover loss maps by @Hansen2013. On the 2000 Harper's map, 200,000 ha clouds are present over the 4.2 million ha moist forest ecoregion (4.8% of clouds). To remove these clouds, we used the 2000 cloud-free tree cover map provided by @Hansen2013 (also at 30m resolution) and we chose a threshold of 75% of tree cover [@Achard2014] to decide whether to replace clouds with forest or non-forest. We thus obtained a cloud-free year 2000 forest-cover map for Madagascar. From this map, using tree cover loss data from 2000 to 2014 provided by @Hansen2013, we obtained annual cloud-free forest-cover maps from 2010 to 2014 at 30m resolution. We corrected the cloudy 1990 Harper forest map assuming that if forest was present in 2000, it was also present in 1990. Doing so, cloud-cover was reduced to a negligeable quantity of `r epsilon_1990`ha. For these residual cloud areas, we assumed that forest-cover was the same as for the year 2000. We also corrected the cloudy 1973 Harper forest map assuming that if forest was present in 1973, it was also present in 1973. Contrary to the year 1990, we were not able to remove all the clouds on the 1973 map and the remaining cloud cover was of `r round(cloud_1973/10e6,2)` million ha. Because the 1953 forest-cover map was obtained from the digitalization of a paper map derived from aerial photographies, and thus cannot be perfectly aligned with the maps obtained in the following years, we did not performed any modifications on the original forest map of 1953 except reprojection. For all forest-cover maps from 1973, we removed the isolated non-forest pixels if all the surrounding pixels were covered by forest, assuming that small-scaled deforestation on a 30m square pixel inside the forest was unlikely. This was important in order not to over-estimate forest fragmentation. All these maps are made freely available at https://bioscenemada.cirad.fr/forestmaps.

### 2.2 Computing forest-cover area and deforestation rate and comparing estimates with previous studies

On the basis of the forest-cover maps we obtained, we computed the forest-cover area (in ha) and the mean annual deforestation rate (in hectares of forest and in percentage) on six decades, for five time periods from _c._ 1953 to 2014 (_c._ 1953-_c._ 1973-1990-2000-2010-2014). Mean annual deforestation rates in percentage were computed as follow [@Puyravaud]: $\theta = 100 \times [1-(1-(F_{t2}-F_{t1})/F_{t1})^{(1/(t2-t1))}]$, $\theta$ being the annual deforestation rate (in %/yr), $F_{t2}$ and $F_{t1}$ being the forest covers (reported in Tab. 2) at times $t2$ and $t1$ (in years). Because of the large cloud-cover in 1973, the annual deforestation rates in hectares of forest for the two periods _c._ 1953-1973 and 1973-1990 are only rough estimates of the deforestation rates. This problem was overcome when computing the annual deforestation rates in percentage for these two periods. In this case, we only considered the cover change and computed the ratio $F_{t2}-F_{t1})/F_{t1}$ for pixels that were not covered by clouds in 1973. Computation were done at the national scale and for each forest type in Madagascar: the moist forest in the East, the dry forest in the West, the spiny forest in the South and mangroves on the West coast. To define the forest type, we used a map provided by the MEFT (_"Ministère de l'Environnement et des Forêts à Madagascar"_) showing the boundaries of the four ecoregions in Madagascar. Ecoregions were defined on the basis of climatic and vegetation criteria using the climate classification by @Cornet1974 and the vegetation classification from the 1996 IEFN national forest inventory [@IEFN1996].

We compared our estimates of forest-cover and forest-cover change with three previous studies that have estimated forest-cover and forest-cover change at the national scale for Madagascar [@Harper2007; @MEFT2009; @ONE2015]. @Harper2007 provided estimates for the periods _c._ 1953-_c._ 1973-1990-2000, @MEFT2009 provided estimates for the periods 1990-2000-2005 and @ONE2015 provided estimates for the periods 2005-2010-2013. The same ecoregion boundaries as in our study have been used in the @ONE2015 report but this was not the case for the @Harper2007 and @MEFT2009 reports, which can explain some of the differences between studies in the forest-cover areas and deforestation rates by ecoregions. W

@@

We report an analysis of the evolution of forest fragmentation based on the forest-cover maps for Madagascar for the years _c._ 1953, _c._ 1973, 1990, 2000, 2010 and 2014. We used the method developed by @Riitters2000 which uses a moving window to characterize the fragmentation around each forested pixel. Computation were done using the function `r.forestfrag` of the GRASS GIS software [@Neteler2008]. Six categories of fragmentation are identified from the amount of forest and its occurrence as adjacent forest pixels: "interior", "perforated", "edge", "transitional", "patch", and "undetermined". We used a moving window of 7x7 pixels (4.41 km<sup>2</sup>). Using this window size, forest edge had a width of about 90 m [@Riitters2000]. The interior category can be interpreted as the most intact forest (sensus @Potapov2017). We reported the area of forest in each fragmentation category for the six years and analysed the evolution of the fragmentation on the six decades. We also computed the distance to forest edge (in km) for all forest pixels for the years _c._ 1953, _c._ 1973, 1990, 2000, 2010 and 2014. For that, we used the function `gdal_proximity.py` of the GDAL software (http://www.gdal.org/). We computed the mean and 95% quantiles of the distance to forest edge and looked at the evolution of these values with time.

@@

@@

```{r defor_comp, results="asis"}
tab.3 <- defor.comp
tab.3$ForestType <- c("Total",rep("",3),"Moist",rep(" ",3),"Dry",rep(" ",3),"Spiny",rep(" ",3),"Mangroves",rep(" ",3))
col.names <- c("Forest type","Source","1953-1973","1973-1990","1990-2000","2000-2005","2005-2010","2010-2013")
kable(tab.3,col.names=col.names,align=c("l","l",rep("r",6)))
```

@@

![Figure 2: **Evolution of the mean distance to forest edge from 1953 to 2014 in Madagascar.** Black dots represent the mean distance to forest edge for each year. Vertical dashed segments represent the 95% quantiles of the distance to forest edge. We observed a continuous decrease of the mean distance to forest edge from 1953 to 2014. The mean distance was of about 1km in 1973 while it was of about 300m in 2014. Moreover, many areas were located far from forest edge (>2km and up to 4km) in 1973, while the majority of the forest was at a distance <1.2 km of the forest edge in 2014.](outputs/dist.png)
